---
title: "Basic Processing of SCov1 and SCov2 infected cells"
output: html_document
  params:
    # minimal number of genes per cell
    n_sub: 2000
    
    # PCA dimensions to use for umap 
    dims_use: 20
---
  
```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo       = FALSE,
  fig.width  = 5, 
  fig.height = 4.5
)
```

# README

This document is used for preliminary processing of single cell data for:

The data can be downloaded from the following Zenodo id: 10.5281/zenodo.4031204

To run the functions, download the data and replace the variable **rds_input_file**

The same processing pipeline can be used for both Calu3 and H1299 cells.

In short, the pipeline installs the dependencies, does scRNA seq pre-processing, 
embedds the data based on diffusion maps, and calculates the velocity field.

## Accessory Functions

```{r Libraries}
# Checks whether the dependencies are installed and loads them
# If they are not installed, installs the dependencies
LoadPackages = function(packages){
  
  # Checks whether BiocManager package exists
  if(!'BiocManager' %in% rownames(installed.packages()))
    stop('Please install the BiocManager package')
  
  # Installs the non existing packages
  non_existing_packages = setdiff(packages, rownames(installed.packages()))
  if(length(non_existing_packages) > 0){
    
    # check for seurat wrappers
    if('SeuratWrappers' %in% non_existing_packages)
       non_existing_packages = c(setdiff(non_existing_packages, 'SeuratWrappers'),
                                 'satijalab/seurat-wrappers')
    
    message('Installing packages ...')
    BiocManager::install(non_existing_packages)
  }
  
  # Loads all of the packages
  message('Loading packages ...')
  invisible(
    lapply(packages, function(x){
      suppressPackageStartupMessages(
        library(x, character.only = TRUE)
      )
    })
  )
}

# ---------------------------------------------------------------------------- #
hh = function(x)x[1:5,1:5]

# ---------------------------------------------------------------------------- #
# uses graph based clustering on diffusion coordinates
Bin_Diffusion_Maps = function(
    lf_mat,
    k_vec = c(3, 5, 10, 25, 50, 100)
){
    message('Cluster louvain ...')
    suppressPackageStartupMessages({
        library(igraph)
        library(FNN)
    })
    lk = list()
    for(k in k_vec){
        message(k)
       knn.norm = get.knn(lf_mat, k = k)
       knn.norm = data.frame(
           from   = rep(1:nrow(knn.norm$nn.index),
           k),
           to     = as.vector(knn.norm$nn.index),
           weight = 1/(1 + as.vector(knn.norm$nn.dist)))
       nw.norm = graph_from_data_frame(knn.norm, directed = FALSE)
       nw.norm = simplify(nw.norm)
       lc.norm = cluster_louvain(nw.norm)
       lk[[paste0('luv_',k)]] = as.factor(membership(lc.norm))
    }
    dk = as.data.frame(lk)
    return(dk)
}

```


## Libraries


```{r, dependson = 'Libraries' }
packages = c(
  "Seurat",
  "dplyr",
  "ggplot2",
  "ComplexHeatmap",
  "igraph",
  "FNN",
  "RColorBrewer",
  "stringr",
  "cowplot",
  "patchwork",
  "destiny",
  "SingleCellExperiment",
  "SeuratWrappers"
)
LoadPackages(packages)
```

# Processing

## Load Data

```{r}
# Input RDS file 
rds_input_file = ''
  
# Here specify the output directory
outpath = './'

```

```{r}
output_directory = file.path(outpath, paste('Velo',n_sub,sep='_'));dir.create(outpath, showWarnings=FALSE)
```


## Preprocessing

```{r}
# read the input seurat object
seu     = readRDS(rds_input_file)

# filter the number of genes
seu_sub = seu[, seu$nFeature_RNA > params$n_sub]

# normalize
seu_sub = NormalizeData(seu_sub)

# scale
seu_sub = ScaleData(seu_sub)

# Find variable features
seu_sub = FindVariableFeatures(seu_sub)

# Run PCA
seu_sub = RunPCA(seu_sub, verbose=FALSE)

# Run UMAP on selected dims
seu_sub = RunUMAP(seu_sub, dims = 1:params$dims_use)
```

## Viral Load Calculation

```{r}
# Extract viral genes
cind1 = grep('^SCoV1', rownames(seu_sub), value=TRUE)
cind2 = grep('^SCoV2', rownames(seu_sub), value=TRUE)

# Extract viral expression
mat  = GetAssayData(seu_sub, 'data')
mat_sel1 = as.matrix(mat[cind1,])
mat_sel2 = as.matrix(mat[cind2,])

# Calculate the load
seu_sub@meta.data$SCov1_Load = colSums(mat_sel1)
seu_sub@meta.data$SCov2_Load = colSums(mat_sel2)
```


## Diffusion Maps

```{r}
# Preapre dataset for diffustion map estimation
feats = VariableFeatures(seu_sub)
DefaultAssay(seu_sub) = 'RNA'
mat   = GetAssayData(seu_sub)[feats,]
mat   = as.matrix(mat)

# Run diffusion map
mat       = mat[!duplicated(mat),]
sce       = SingleCellExperiment(assays = list(logcounts = mat))
dm_rna    = DiffusionMap(sce)

# Run UMAP on diffusion map embeddings
umap_rna  = RunUMAP(eigenvectors(dm_rna), reduction.key = 'UMAPrna_')
rownames(umap_rna@cell.embeddings) = colnames(seu_sub)
seu_sub$umap_rna = umap_rna

# Fetch the diffusion map KNN
dm_dpt                = DPT(dm_rna)
dm_dist_mat           = as.matrix(dm_dpt)
rownames(dm_dist_mat) = colnames(dm_dist_mat) = colnames(mat)

# Run diffusion map based knn binning
dm_clust = Bin_Diffusion_Maps(eigenvectors(dm_rna))

dm_ev = data.frame(cell_id = colnames(seu_sub), eigenvectors(dm_rna))

colnames(dm_clust) = paste0('virus_',colnames(dm_clust))
seu_sub@meta.data = cbind(seu_sub@meta.data, dm_clust)
```


## Velocity Estimation

```{r}
# runs the velocity estimation 
velo = RunVelocity(
  seu_sub,
  spliced           = "RNA",
  unspliced         = "Unspliced",
  ambiguous         = NULL,
  spliced.average   = 0.1,
  unspliced.average = 0.05,
  reduction         = "pca",
  group.by          = "ident",
  deltaT            = 1,
  kCells            = 50,
  fit.quantile      = 0.05,
  ncores = 16, verbose = TRUE
)

delta    = as.matrix(Tool(object = velo, slot = 'RunVelocity')$deltaE)
seu_sub$Velo = CreateAssayObject(data = delta)
```


# Save the resulting Seurat object

```{r}
outfile = file.path(output_directory, 'Seurat_Processed.rds')
saveRDS(seu_sub, outfile)
```

